# Creating Effective Data Visualizations (in R)

### Case Western Reserve University Journal of Economics

For this module the ggplot package in R will be used. In the R console type `install.packages('ggplot2')`. Alternatively, the tidyverse package contains ggplot2, and also comes with packages such as dplyr that can be used for data cleaning and processing. To install this instead of ggplot2 you can type in the R console `install.packages('tidyverse')`

**Loading ggplot2**\
In order to load ggplot2 into your R or R-markdown file, the following command must be included in your script or document. (Important to note: This must be inputted prior to using any of the packages)

`library(ggplot2)`

### Loading Datasets

In R we will load our datasets with different commands based on the format of our dataset.

**CSV**

This is the most common format of dataset we will come across. CSV stands for comma-separated values. In R we can use the following command to load the dataset:\
`var_name <- read.csv('filepath')` This will set the var_name to the data, and will load the data from the file path (this can be an absolute or relative path (relative means from the folder your file is in, absolute would be from the root directory of your operating system))

If you do not have headers (this is a fairly rare case) then u must include the headers parameter (it is by default True):\
`var_name <- read.csv('filepath', header=F)`

**XLSX (Microsoft Excel)**

This is another common format of dataset. In R however we need another library in order to read them in well.

In the console install `readxl` using `packages.install('readxl')`. Then in your document use `library(readxl)`

Then we can use: `var_name <- read_excel('filepath')`\
There are lots of powerful optional arguments that you can refer to the documentation on. The only one that could be generally useful is the sheet_name parameter:

Unlike Python, R is a 1 index language, so it is defaulted at 1 (so the first sheet). If you need to pull data from a different sheet the simplest way to do this is to specify the sheet number or name in the sheet_name parameter.

Imagine I have a excel sheet with 2 sheets, original_data, and corrected_data. If I use the command with just the filepath I will pull the data from the original_data sheet. If I want the corrected_data, I have 2 options:

-   Use the index:

`var_name <- read_excel('filepath', sheet=2)`

-   Use the sheet name:

`var_name <- read_excel('filepath', sheet='corrected_data')`

**TXT**

These are much less common but some old census files I have seen use pipe-delimited text files. In this case you will have to use a different command and declare the delimiter if it is anything but tab delimited.

`var_name = read.table('filepath', sep = '|')`

\^ This example is for pipe-delimited text file parsing.

The data would look something like this:

> name \| age \| income\
> Mark \| 25 \| 0

**RData**

This is a data format that is unique to R, RData files are not frequently available from most data sources, but you can saved clean data as RData files, and you could run into them as R users. These are the easiest to load:

`var_name <- load('filepath')`

\*I would not recommend using this format of data to save your cleaned data if your teammate uses Python, in general csv is the easiest to work with across languages and would be my recommendation for data formats.

For any other formats of data (JSON, XML, SHP, etc) reach out to your project leads or me for help (often ChatGPT can help)

```{r}
## For this example, we will be sourcing our data from the MTA Daily Ridership Data.
##      In this cell block we will read in our data and look at the columns that we have
##      as well as what our data looks like.

##  As we can see, we have a timeseries of data that is indexed daily,
##      our columns are split by method of transit and for each method of transit we have
##      the number of riders for that day as well as the % of ridership compared a comparable pre-pandemic day
## TO use thse function you will need to do `install_packages()` for tidyverse and lubridate
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidyverse) 
data <- read.csv('MTA_Daily_Ridership_Data__Beginning_2020_20240629.csv')
print(names(data))
head(data)
## R peculiarity so we will rename the columns 

data <- data %>%
  rename(
    Date = `Date`,
    `Subways: Total Estimated Ridership` = `Subways..Total.Estimated.Ridership`,
    `Subways: % of Comparable Pre-Pandemic Day` = `Subways....of.Comparable.Pre.Pandemic.Day`,
    `Buses: Total Estimated Ridership` = `Buses..Total.Estimated.Ridership`,
    `Buses: % of Comparable Pre-Pandemic Day` = `Buses....of.Comparable.Pre.Pandemic.Day`,
    `LIRR: Total Estimated Ridership` = `LIRR..Total.Estimated.Ridership`,
    `LIRR: % of Comparable Pre-Pandemic Day` = `LIRR....of.Comparable.Pre.Pandemic.Day`,
    `Metro-North: Total Estimated Ridership` = `Metro.North..Total.Estimated.Ridership`,
    `Metro-North: % of Comparable Pre-Pandemic Day` = `Metro.North....of.Comparable.Pre.Pandemic.Day`,
    `Access-A-Ride: Total Scheduled Trips` = `Access.A.Ride..Total.Scheduled.Trips`,
    `Access-A-Ride: % of Comparable Pre-Pandemic Day` = `Access.A.Ride....of.Comparable.Pre.Pandemic.Day`,
    `Bridges and Tunnels: Total Traffic` = `Bridges.and.Tunnels..Total.Traffic`,
    `Bridges and Tunnels: % of Comparable Pre-Pandemic Day` = `Bridges.and.Tunnels....of.Comparable.Pre.Pandemic.Day`,
    `Staten Island Railway: Total Estimated Ridership` = `Staten.Island.Railway..Total.Estimated.Ridership`,
    `Staten Island Railway: % of Comparable Pre-Pandemic Day` = `Staten.Island.Railway....of.Comparable.Pre.Pandemic.Day`
  )
```

### Effective Line Graph

Effective line graphs avoid using too many lines, and too many colors. It is key to keep graphs simple, our line graphs are to help the reader quickly understand our argument.

In this portion we will learn how to use the python library `ggplot2` to do create a line graph.

In `ggplot2` we can set our color using the `color` parameter. We can either supply a hexcode or a string like 'blue'. We should have Journal theme hexcodes available by the time we start publishing. In `ggplot2` we add lines to a plot using `geom_line` and we use `aes` to specify our y or our x.

```{r, fig.width=15, fig.height=5}

data <- data %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))  

# Create the plot
ggplot(data, aes(x = Date)) +
  geom_line(aes(y = `Subways: Total Estimated Ridership`, color = "Subway Ridership")) +
  geom_line(aes(y = `Bridges and Tunnels: Total Traffic`, color = "Bridge and Tunnel Traffic")) +
  scale_color_manual(values = c("Subway Ridership" = "#0B4F6C", "Bridge and Tunnel Traffic" = "#9197AE")) +
  labs(title = "Subway Ridership vs. Bridge and Tunnel Traffic in NYC (2020-2024)",
       x = "Year",
       y = "Number of Persons/Vehicles",
       color = "Legend") +
  theme_minimal() +
  theme(legend.position = "bottom")

```
This graph above is ok, but its messy, the data is too frequent for this long period of time, lets try and graph by week instead. 
```{r, fig.width=15, fig.height=5}

# Step 1: Remove NA values in Date and other columns
data <- data %>% filter(!is.na(Date))

# Step 2: Create Year and Week columns
data <- data %>%
  mutate(
    Year = year(Date),        # Extract year
    Week = floor_date(Date, "week")  # Use floor_date to get the first day of the week
  )


# Step 3: Drop unnecessary columns
data <- data %>%
  select(-Year)

# Step 4: Group by Week and calculate mean
data_grouped <- data %>%
  group_by(Week) %>%
  summarise(across(everything(), mean, na.rm = TRUE), .groups = 'drop')


ggplot(data_grouped, aes(x = Week)) +
  geom_line(aes(y = `Subways: Total Estimated Ridership`, color = "Subway Ridership"), size = 1) +
  geom_line(aes(y = `Bridges and Tunnels: Total Traffic`, color = "Bridge and Tunnel Traffic"), size = 1) +
  scale_color_manual(values = c("Subway Ridership" = "#0B4F6C", "Bridge and Tunnel Traffic" = "#9197AE")) +
  labs(title = "Subway Ridership vs. Bridge and Tunnel Traffic in NYC (2020-2024)",
       x = "Week",
       y = "Number of Persons/Vehicles",
       color = "Legend") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
### Effective Scatter Plots

Scatter plots are great at showing the relationships between two variables. We can also hightlight certain datapoints that we want to hightlight, maybe it corresponds to a particular area of interest (like below highlighting during Covid) Or maybe we are identifying interesting outliers. The goal of a scatter plot is to show some sort of trend in the data between two variables. To make scatter plots we simply use `plt.scatter(x, y)` `label` and `color` are still optional parameters that we can set.

```{r, fig.width=10, fig.height=5}
# Step 1: Create a new column to indicate if the date is during COVID
data <- data %>%
  mutate(CovidHighlight = ifelse(Date > as.Date("2020-03-19") & Date < as.Date("2021-06-01"), "Covid-19 Dates", "Non Covid Days"))

# Step 2: Create the scatter plot
ggplot(data, aes(x = `Bridges and Tunnels: Total Traffic`, y = `LIRR: Total Estimated Ridership`, color = CovidHighlight)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Non Covid Days" = "#9197AE", "Covid-19 Dates" = "#0B4F6C")) +
  labs(title = "LIRR Ridership vs. Bridge and Tunnel Traffic in NYC (2020-2024)",
       x = "Number of vehicles",
       y = "Number of persons",
       color = "Legend") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_continuous(labels = scales::comma) +  
  scale_y_continuous(labels = scales::comma) +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Effective Bar Graphs

Bar charts are excellent for comparing quantities across different categories. They provide a clear visual representation of data, making it easy to see which categories are larger or smaller. You can also highlight specific bars to draw attention to particular data points, such as those that are above a certain threshold or represent significant changes.

To create a bar chart in ggplot2, you can use `geom_barbar(stat, position)`, where x represents the categories and height represents the values. You can customize the appearance using optional parameters like `color`. 

```{r, fig.width=12, fig.height=6}
# Step 1: Filter and summarize the data
year_data <- data %>%
  filter(year(Week) %in% c(2020, 2021, 2022)) %>%
  mutate(year = year(Week)) %>%
  group_by(year) %>%
  summarise(
    `Subways: Total Estimated Ridership` = sum(`Subways: Total Estimated Ridership`, na.rm = TRUE),
    `Buses: Total Estimated Ridership` = sum(`Buses: Total Estimated Ridership`, na.rm = TRUE),
    `LIRR: Total Estimated Ridership` = sum(`LIRR: Total Estimated Ridership`, na.rm = TRUE),
    `Metro-North: Total Estimated Ridership` = sum(`Metro-North: Total Estimated Ridership`, na.rm = TRUE),
    `Bridges and Tunnels: Total Traffic` = sum(`Bridges and Tunnels: Total Traffic`, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -year, names_to = "Transportation Mode", values_to = "Estimated Ridership")

# Step 2: Define colors
colors <- c("#00008B", "#1E90FF", "#00CED1", "#A0C4E0", "#4682B4")

# Step 3: Create the bar plot
ggplot(year_data, aes(x = as.factor(year), y = `Estimated Ridership`, fill = `Transportation Mode`)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors) +
  labs(title = 'Average Ridership by Transportation Mode (2020-2022)',
       x = 'Year',
       y = 'Average Estimated Ridership') +
  scale_y_continuous(labels = scales::comma) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5), legend.title = element_blank())

# Step 4: Create the log scale plot
ggplot(year_data, aes(x = as.factor(year), y = `Estimated Ridership`, fill = `Transportation Mode`)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors) +
  labs(title = 'Average Ridership by Transportation Mode (2020-2022) on Log Scale',
       x = 'Year',
       y = 'Average Estimated Ridership') +
  scale_y_log10(labels = scales::comma) +  # Log scale y-axis
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5), legend.title = element_blank())
```
### Effective Histograms

Histograms are useful for displaying the distribution of a dataset, showing how frequently different ranges of values occur. They help identify patterns, such as skewness, modality, and the presence of outliers. Highlighting specific bins can emphasize certain ranges of interest, like values that fall within a target range or outliers.

To create a histogram in ggplot2, you can use `geom_histogram(bins, fill, color)`, where data is the dataset and bins defines how many intervals you want to divide your data into.
For example, below we see a bimodal distribution of sorts on Subway Ridership
```{r, fig.width=10, fig.height=5}

ggplot(daily_data, aes(x = `Subways: Total Estimated Ridership`)) +
  geom_histogram(bins = 30, fill = "#00008B", color = "black", alpha = 0.75) +
  labs(title = "Distribution of Daily Subway Ridership (2020-2024)",
       x = "Number of Riders",
       y = "Frequency") +
  scale_x_continuous(labels = scales::comma)
  theme_minimal() +
  theme(panel.grid.major.y = element_line(), 
        panel.grid.minor = element_blank())  
```

